OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-11-19 manuscript_nar.tex
adding GM results and minor typos

\author{%
Enrique Vidal\,$^{1,2,*}$,
Fran\c{c}ois le Dily\,$^{1,2}$,
Javier Quilez\,$^{1,2}$,
Ralph Stadhouders\,$^{1,2}$,
Yasmina Cuartero\,$^{1,2,3}$,
Thomas Graf\,$^{1,2}$,
Marc A.  Mart\'i-Renom\,$^{1,2,3,4}$,
Miguel Beato\,$^{1,2}$,
and Guillaume J. Filion\,$^{1,2}$%
\footnote{To whom correspondence should be addressed.
Tel: +34 93 316 01 15; Fax: +34 93 316 00 99; Email: enrique.vidal@crg.eu}}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
contribution of known biases via a generalized additive model. The
normalized data is more reproducible between replicates and across
different protocols. Importantly, \textit{OneD} is also efficient when
cells have a normal karyotype, where it performs as well as the best
normalization methods. Finally, the implementation is as fast as
\textit{ICE} and it scales up to 1~kbp resolution with reasonable
computing resources.

@@

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\lambda_i'$). The model consists of 8 states that correspond to 1, 2, 3,
4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
emission parameters (a single position parameter for states 1-7, a
position parameter for state `8 or more', a single standard deviation for
all the states and a single degree of freedom for all the states) and 56
transition parameters.

@@

We used several experiments comprising T47D breast cancer cell lines (7
samples), K562 leukemia cell lines (4 samples), both with aberrant
karyotypes; and mouse primary B cells (6 samples ) and ES cells (7
samples), both with normal diploid karyotypes. The experiments were
carried out in different laboratories, following either the original Hi-C
protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
version \citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MboI and NcoI). We also used array-based copy-number
segmentation of the two cell lines obtained from the COSMIC database
\citep{forbes2010cosmic}.

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~kbp, but we found no major differences at different
resolutions (not shown).

@@

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Supplementary
Table~1 for details). Given a set of experiments and a metric, we first
computed all pairwise combinations between experiments and then classified
the comparisons according to the characteristics of each pair (cell type,
protocol, batch and treatment).

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were computed
using the \texttt{ROCR} package \citep{sing2005rocr}.

@@

In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
number correction (\textit{OneD+CN}) against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
servant2012hitc}. The first four methods correct biases implicitly,
whereas the fourth method does it explicitly.

@@

On the T47D data set, the methods that increased the reproducibility
between identical cell types also increased it between different cell
types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
exclusively the biological signal. However, the difference was most marked
for \textit{OneD} and \textit{caICB}. On the K562 data set
(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
between identical cell types but not between different cell types
Figures~\ref{fig:aberrant}B), which is the desired behavior of a
correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
The conclusions were unchanged when using other metrics to compare Hi-C
matrices (Supplementary Figures~4, 5 and 6).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the larest area
under the curve in both tests, and the corresponding ROC curve was above
the others throughout. Using other metrics to compare Hi-C matrices
yielded similar results (Supplementary Figures~4, 5 and 6).

@@

@@

The results so far suggest that the copy number correction lowers the
performance of \textit{OneD}. A possible limitation in Hi-C is the absence
of ground truth. We thus used simulations to generate matrices with
defined karyotypic aberrations (see Materials and Methods) and tested the
capacity of the correction methods to distinguish different cell types
(Supplementary Figure~11). On this test, \textit{LGF} and \textit{OneD}
performed best. In comparison, \textit{OneD+CN} performed less well,
suggesting that correcting the copy number attenuates the biological
signal. This is corroborated by the fact that the copy number correction
alone has a lower performance than no correction at all. Thus, copy number
correction can indeed have side effects that lower the biological signal.

@@

\textit{OneD} was the third fastest method and it always ran in less than
1~min in the conditions of the benchmark. The copy number correction
increased the running time, but it remained unert 1~min in general.
Throughout this benchmark, the memory footprint of \textit{OneD} was less
than 300 MB.  Interestingly, the running time of \textit{OneD} was much
more homogeneous than that of the other methods. The reason is that the
size of the regression problem to be solved by the \texttt{mgcv} package
is always the same at fixed resolution.

@@

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower resolutions. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balacing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~12). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually the faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution reslution (5~kbp) it appeared only at high sequening
depth.

>>>>

\author{%
Enrique Vidal\,$^{1,2,*}$,
Fran\c{c}ois le Dily\,$^{1,2}$,
Javier Quilez\,$^{1,2}$,
Ralph Stadhouders\,$^{1,2}$,
Yasmina Cuartero\,$^{1,2,3}$,
Thomas Graf\,$^{1,2}$,
Marc A.  Marti-Renom\,$^{1,2,3,4}$,
Miguel Beato\,$^{1,2}$,
and Guillaume J. Filion\,$^{1,2}$%
\footnote{To whom correspondence should be addressed.
Tel: +34 93 316 01 15; Fax: +34 93 316 00 99; Email: enrique.vidal@crg.eu}}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 2~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
contribution of known biases via a generalized additive model. The
normalized data is more reproducible between replicates and across
different protocols. Importantly, \textit{OneD} is also efficient when
cells have a normal karyotype, where it performs as well as the best
normalization methods. Finally, the implementation is as fast as
\textit{ICE} and it scales up to 5~kbp resolution with reasonable
computing resources.

@@

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\hat{t}_i = \sum_j^n{\hat{x}_{ij}}$). The model consists of 8 states that correspond to 1, 2, 3,
4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
emission parameters (a single position parameter for states 1-7, a
position parameter for state `8 or more', a single standard deviation for
all the states and a single degree of freedom for all the states) and 56
transition parameters.

@@

We used several experiments comprising T47D breast cancer cell lines (6
samples), K562 leukemia cell lines (8 samples), both with aberrant
karyotypes; mouse primary B cells (6 samples) and ES cells (7
samples), both with normal diploid karyotypes; GM12878 B-lymphoblastoid cells
(58 samples), with diploid karyotype; BT474, MCF10, MCF-7 and SKBR3 breast cell
lines (1 sample each). The experiments were
carried out in different laboratories, following either the original Hi-C
protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
version \citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MspI, MboI and NcoI). We also used array-based copy-number
segmentation of the two cell lines obtained from the COSMIC database
\citep{forbes2010cosmic}.

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~kbp, but we found no major differences at different
resolutions (not shown).

@@

We defined five data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples, the fourth contained all
GM12878 samples, the fifth contained one sample of each breast cells (see Supplementary
Table~1 for details). Given a set of experiments and a metric, we first
computed all pairwise combinations between experiments and then classified
the comparisons according to the characteristics of each pair (cell type,
protocol, batch and treatment).

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were computed
using the \texttt{ROCR} package \citep{sing2005rocr} for the three first sets
(those including samples of different cell types).

@@

In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
number correction (\textit{OneD+CN}) against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
servant2012hitc}. The first four methods correct biases implicitly,
whereas the fifth method does it explicitly.

@@

On the T47D data set, the methods that increased the reproducibility
between identical cell types also increased it between different cell
types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
exclusively the biological signal. However, the difference was most marked
for \textit{OneD} and \textit{caICB}. On the K562 data set
(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
between identical cell types but not between different cell types
%Figures~\ref{fig:aberrant}B)
, which is the desired behavior of a
correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
The conclusions were unchanged when using other metrics to compare Hi-C
matrices (Supplementary Figures~4, 5 and 6).

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). \textit{OneD} had the largest area
under the curve in both tests, and the corresponding ROC curve was above
the others throughout. Using other metrics to compare Hi-C matrices
yielded similar results (Supplementary Figures~4, 5 and 6).

@@

We took advantage of the wide range of protocols used in the GM12878 set to
further investigate the performance of each method. The differences between
methods were minor and \textit{OneD} showed the greatest improvement in
reporducibility (Supplementary Figure~11).

@@

The results so far suggest that the copy number correction lowers the
performance of \textit{OneD}. A possible limitation in Hi-C is the absence
of ground truth. We thus used simulations to generate matrices with
defined karyotypic aberrations (see Materials and Methods) and tested the
capacity of the correction methods to distinguish different cell types
(Supplementary Figure~12). On this test, \textit{LGF} and \textit{OneD}
performed best. In comparison, \textit{OneD+CN} performed less well,
suggesting that correcting the copy number attenuates the biological
signal. This is corroborated by the fact that the copy number correction
alone has a lower performance than no correction at all. Thus, copy number
correction can indeed have side effects that lower the biological signal.

@@

On average, \textit{OneD} was the second fastest method and it always ran in less than
1~min in the conditions of the benchmark. The copy number correction
increased the running time, but it remained under 1~min in general.
Throughout this benchmark, the memory footprint of \textit{OneD} was less
than 300 MB.  Interestingly, the running time of \textit{OneD} was much
more homogeneous than that of the other methods. The reason is that the
size of the regression problem to be solved by the \texttt{mgcv} package
is always the same at fixed resolution.

@@

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower resolutions. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balacing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution reslution (5~kbp) it appeared only at high sequening
depth.
